---
title: "Final Project"
author: "Lacey Shivers"
date: "12/8/2021"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

The final project covers all the main components for a statistical analysis required by the final project and in report style. All explanation and coding for 
reproducibility of analysis is shown through the project, done in Rmarkdown.

The topic I decided on for my final project is major league baseball (MLB) statistics. The statistical data to be used is collected from the official MLB website and 
is a publicly open source for people to use. The website holds data stats over all the baseball teams and individual league players. The MLB website is updated regular 
with the newest 2021 season for all players and teams. Looking over the data, an initial observation is the site clearly organized the data to be easily accessible. 
Being able to bring together the wide variety of data on each player and team, helps cultivate research questions. In choosing this data, I plan on using the data to 
answer the 'big' statistical question with analytically processing.  

With the MLB data, I am able to come up with a statistical model for analysis of major league baseball teams. Looking over the data for related statistical questions.  

The following project will address a topic of Major League Baseball, laying out the big research question. Overall the project will answer the big question using 
visualizations, statistical analysis, and have a detailed conclusion. The documentation of data will allow for replication of the data clean up processing and 
analytically process. The presentation of the project will be in a readable layout for an easy understanding for an outside reader.

# Baseball data; Gathered data

The below data is obtained and downloaded from MLB Stats website (https://www.mlb.com/stats/). The data downloaded contains data covering the hitting and pitching for 
the 30 MLB teams during the post season. With the post season data for the teams, the batting averages for the teams is included for the 2021 season. All the data 
obtained will be used to make a thorough analysis on the major league baseball teams, in attempt to answer the question concerning the teams. 

```{r Collected data}
data1 <- read.csv("exit_velocity (1).CSV") # team hitting post season
head(data1)

data2 <- read.csv("exit_velocity (2).CSV") # team pitching post season 
head(data2)

data3 <- read.csv("outs_above_average.CSV") # batting team averages of 2021
head(data3)
```
## cleaning / processing data

After collecting the data from the open source of the most recent data, cleaning the data is the next necessary step. The cleaning process shown below is to done by 
using processing of filtering out unfinished data. Cleaning out any empty rows and columns to get the most accurate representation of the data for analytically 
processing.

```{r cleaning data}
library(tidyverse)
library(janitor)

# c = clean 

cdata1 <- data1 %>% remove_empty(whic=c("rows"))
cdata1 <- data1 %>% remove_empty(whic=c("cols"))

cdata2 <- data2 %>% remove_empty(whic=c("rows"))
cdata2 <- data2 %>% remove_empty(whic=c("cols"))

cdata3 <- data3 %>% remove_empty(whic=c("rows"))
cdata3 <- data3 %>% remove_empty(whic=c("cols"))

```

The cleaned data above has removed any columns or rows with missing information for each dataset.

## Framing a statistical question with the cleaned MLB data

The cleaned MLB stats data is able to help frame potential answerable questions, in regards to the major league baseball teams. This dataset is reasonable because it 
pertains strictly to the MLB teams stats. The beginning of the analysis brings on the problem and question of hitters, pitchers and batting team averages.

Do hitters, pitchers, or batting averages of teams have a better relation to the distance and speed of the angle of hits or outs by batting of the baseball teams as a 
whole? This question will help answer the problem of the most significance for hits by team hitters/ batters and pitchers. The gathered and cleaned datasets are good 
options for answering the proposed question, because it covers all the aspects of the question. 

## statistical analysis

Analysis for potentially answering the statistical question. Using course learning materials for implementation.

## visualizations 

Visualizations to get an idea of the presented data. Seeing the visualizations to represent the beginning of the analysis. 

```{r visualizations}
library(GGally) 

ggcorr(cdata1) # correlation plot of cleaned data1

# sample column of strong positive correlated data
hist(cdata1$avg_hit_speed,freq=FALSE)

qqnorm(cdata1$avg_hit_speed)
qqline(cdata1$avg_hit_speed)

ggcorr(cdata2) # correlation plot of cleaned data2

# sample column of strong positive correlated data
hist(cdata2$ev95plus,freq=FALSE)

qqnorm(cdata2$ev95plus)
qqline(cdata2$ev95plus)

ggcorr(cdata3) # correlation plot of cleaned data1

# sample column of strong positive correlated data
hist(cdata3$outs_above_average,freq=FALSE)

qqnorm(cdata3$outs_above_average)
qqline(cdata3$outs_above_average)

```
The correlations of each dataset shown above allows us to see the relation between all the variables. Using a strongly correlated variable to see if it is compatible 
against normalcy. As all the chosen variables shown, they have a normal distribution within the datasets. With the correlation plots the columns of data for hitters 
and pitchers for the post season show a mostly strong positive correlation. As for the batters 2021 average dataset correlation it shows a low correlation between the 
column variables. Having this information allows us to move into a deeper analysis of the stronger correlated variables for each of the datasets.

## analysis continues

```{r analysis}

# summary analysis of the cleaned datasets
summary(cdata1)
summary(cdata2)
summary(cdata3)

library("ggpubr")

## multiple linear regression 

# dataset1 hitting
fit1 = lm(avg_hit_angle~avg_hit_speed+avg_distance+avg_hr_distance, data= cdata1)
summary(fit1)

ggscatter(cdata1,x = "avg_hit_angle", y = "avg_distance", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson")

#dataset2 pitchers # using the same varaibles as dataset1 to see a comparison between hitters and pitchers of the MLB teams
fit2 = lm(avg_hit_angle~avg_hit_speed+avg_distance+avg_hr_distance, data= cdata2)
summary(fit2)

ggscatter(cdata2,x = "avg_hit_angle", y = "avg_distance", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson")

#dataset3 batting averages 
fit3 = lm(outs_above_average~outs_above_average_rhh+outs_above_average_behind+outs_above_average_infront, data= cdata3)
summary(fit3)

ggscatter(cdata3,x = "outs_above_average", y = "outs_above_average_rhh", 
          add = "reg.line", conf.int = TRUE, 
          cor.coef = TRUE, cor.method = "pearson")

```
The multiple linear regression shows a relationship between strong positive correlated average variables of the datasets. Using the models for predicting the value of 
avg_hit_angle based on avg_hit_speed, avg_distance, and avg_hr_distance. The summary of the multiple linear fits shown above, show how significant each coefficient is 
for each dataset of the teams. Using a scatterplot to see the most significant coefficient for the correlation between the chosen data. 

# analysis continues

Checking the residuals of the linear models to see if multiple linear regression is a good analysis for the major league baseball datasets. The summary of fits for the 
multiple linear regression shows the coefficients and R squared outputs. For a more clear representation of the outputs, futher analysis will be shown. Looking at the 
determination of coefficients for each dataset model fit and the residuals to continue to check for a normal distribution.  
```{r coefficients}
# coefficient of determination for fit1
coefficients(fit1)
summary(fit1)$r.squared

# plot of the residuals to check if the residuals are normally distributed
f.res1 = resid(fit1)
qqnorm(f.res1, main ="Residuals check") 
qqline(f.res1)

# coefficient of determination for fit2
coefficients(fit2)
summary(fit2)$r.squared

# plot of the residuals to check if the residuals are normally distributed
f.res2 = resid(fit2)
qqnorm(f.res2, main ="Residuals check") 
qqline(f.res2)

# coefficient of determination for fit1
coefficients(fit3)
summary(fit3)$r.squared

# plot of the residuals to check if the residuals are normally distributed
f.res3 = resid(fit3)
qqnorm(f.res3, main ="Residuals check") 
qqline(f.res3)

```

From the above residuals analysis, there is a trend for each of the model fits. With only 1 to 3 outliers and a good trend the results are fair for a continued 
analysis. 

# analysis continue

The continued analysis using the strong significant coefficients from the multiple linear regression for correlations between the found individual dataset correlation.

```{r continue}

# correlations picked from the significance of the coefficients 
d = cdata1$avg_distance # distance of hits done by team hitters
f = cdata1$avg_hit_angle # angle of hit done by team hitters
cor(d,f) # strong positive correlation

g = cdata2$avg_distance # distance of hits done by team pitchers
h = cdata2$avg_hit_angle # angle of hit done by team pitchers
cor(g,h) # highly positive correlation

# checking correlations between the teams for best averages
ggplot(cdata1, aes(x=avg_hit_angle, y=avg_distance, color= ï..team)) + geom_point()

ggplot(cdata2, aes(x=avg_hit_angle, y=avg_distance, color= ï..team)) + geom_point()


# comparing datasets, seeing a better pitching or hitting correlation of the teams. 
ggplot() +
  geom_point(data = cdata1, aes(x = avg_hit_angle, y = avg_distance), color= "blue") +
  geom_point(data = cdata2, aes(x = avg_hit_angle, y = avg_distance)) + geom_abline(intercept = 104, slope = 5)

```
The correlations of the chosen variables show the data continues to be correlated. With the correlations the above scatterplot visualizations show individual team 
correlations between the average hitting distance and the average hit angle. The colored teams for the first two scatterplots show which top teams are more accurate 
for hitting and pitching. The last visualization shows the comparison of the two teams averages overall. Showing that the hitting teams averages are more accurate than 
the average pitchers teams. 

# analysis continues
 
With similar analysis' outcomes of average hit angles for pitchers and hitters of the major league baseball teams, I decided to combine the first two datasets. The 
combined dataset will help further the understanding for a more in depth answer concerning the big question. The continued analysis allows for a more thorough and 
precise analysis. 

```{r continue 2} 
defaultW <- getOption("warn") # gets rid of warning
options(warn = -1) 

totaldata <- rbind(cdata1,cdata2) # combines the pitching and hitting teams data

plot(totaldata[,6:9]) # linear pairs 

library(ISLR)
library(MASS)
set.seed(1)

# Linear Discriminant Analysis (LDA)
lda.fit=lda(avg_hit_speed ~ fbld + gb + max_hit_speed ,data=totaldata)
lda.fit

# Assessing accuracy of a classifier
lda.pred=predict(lda.fit)
names(lda.pred)

predicted.class=lda.pred$class
predicted.class

# Confusion matrix
table(predicted.class[1:10],totaldata$avg_hit_speed[1:10])

# 5-fold cross validation with MSE
library(boot)
library(splines)

mse <- rep(NA, 5)
for (i in 1:5) {
    fit <- glm(totaldata$avg_hit_speed ~ ns(totaldata$fbld,i) + totaldata$gb + totaldata$max_hit_speed, data= totaldata)
    deltas <- cv.glm(totaldata, fit)$delta[1]
    
    mse[i] = mean((deltas - predict(fit, newdata = list(dis=totaldata$avg_hit_speed)))^2)
}
mse
plot(mse, type = "l")

# checking the mean mse
ns.fit <- lm(totaldata$avg_hit_speed ~ ns(totaldata$fbld,i) + totaldata$gb + totaldata$max_hit_speed, data= totaldata)

x <- summary(ns.fit)

mean(x$residuals^2)

```
The continued analysis shown above is the linear discriminant analysis to show how to the combined totaldata is to find a linear combination of features that 
characterizes the classes of objects in the data. Using a prediction function to predict the values of the LDA to further check the analysis of the data. checking the 
prediction of the LDA model with a confusion matrix, brings more clarity. The confusion matrix represents that the prediction was fairly accurate, meaning our 
totaldata is a great representation for avg_hit_speed, fbld and gb, and max_hit_speed. From the LDA analysis, using natural splines, I checked for the best fitting 
degree using a 5 k-fold cross validation and computing the mse. The technical output for the analysis is degree 5 being the lowest mse of the totaldata. 

This connects with the beginning analysis because both prove and connect back to the big question. The research answers that there is a relation between the major 
league baseball stats with reference to the variables. The totaldata concurred that the hitting teams averages are more accurate than the average pitchers teams over 
the batting outs of the 30 teams as a whole. 

## Conclusion

The project covers a range of analysis that is justified and backed up with reasonable explanation. Understanding that the gathered data has normalcy and are 
correlated among the variables. This information allowed for a further analysis of multiple linear regression (MLR) and additional analysis with linear discriminant 
analysis (LDA). 

I selected the methodology by seeing and checking what is the best fit for representing the gathered and cleaned data. The chosen methods mentioned have pros and cons 
but overall are the necessary methods for presenting the data. The MLR is a pro for being able to determine the relative influence for more than one predictor 
variables and has the ability to identify outliers. This is useful for correlation purposes. A disadvantage to using the MLR method would be if the data was 
incomplete, but because of the cleaning process and the columns and rows are filled with complete baseball stats data. As for the LDA method an advantage or pro for 
this method is that it uses information from both the predictors of totaldata to create a new axis, then in turn minimizes the variance of the multiple variables. As 
for a con of LDA it requires normal distribution assumption on predictors. However this is not an issue because it was proven/ shown with analysis that the MLB stats 
data is normally distributed, described in the beginning of the final project analysis. 

With the covered analysis for the data, I do not think additional analysis is required for the statistical question presented. Although for a broader research on major 
league baseball stats, individual team player stats could be collected. Collecting this data allows for more individual analysis for players instead of teams. This 
extension of analysis opens up more aspects to be looked into for different statistical questions. I do not think this individual player data is necessary for this 
final project on teams in the major league baseball.  

In conclusion, the statistical analysis covers how the major league baseball teams as a whole compared in different analysis'. The completed analysis allows for a 
prediction answer for the statistical problem question of hitting and pitching concerning teams stats. The MLB stats analysis addresses the approach of the question 
thoroughly in context and explanation. The analysis through out the project is explained in each part by beginning with visualization of the MLB stats data, following 
with multiple linear regression (MLR) and ending with a 5-fold cross validation technique. The depth of the chosen methods is to further evaluate the MLB data from MLR 
to using a LDA checked with a 5-fold cross validation technique for the mse. These outcomes help support the prediction of an answer for the thought of statistical 
question. The question of, 'Do hitters, pitchers, or batting averages of teams have a better relation to the distance and speed of the angle of hits or outs by batting 
of the baseball teams as a whole?'. From the evaluation of all the Major League Baseball stats data, it is shown that the hitters of the MLB teams averages when 
related to distance and speed by angle of hits are more significant than the other corresponding aspects. This is also represented in the batters average outs in the 
data, because they are less related between the outs by hitters. Overall, the depth of analysis was sufficient for this statistical process. Covering from gathering 
and cleaning the data to providing evidence of analysis for a complete project of major league baseball stats. 
